---
title: "ToxPi Sample Clustering"
author: "Melanie Odenkirk"
output:
  html_document:
    theme: united 
---

Example dataset for mapping metadata to sample profile clustering. Patient profiles of the top 20 significant lipids are uploaded into ToxPi software following a variance transformation to correct for lipid variability across species. Results are then uploaded here, scaling max/min values are removed, and clinical information in the form of categorical variables (Other infections) and numerical variables (Log(sample collection time relative to delivery)) are mapped onto sample clustering. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", tidy.opts = list(width.cutoff = 60), tidy = TRUE, echo = FALSE, include = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = 'data/')

  library(httr)
  library(jsonlite)
  library(fingerprint)
  library(rcdk)
  library(ggplot2)
  library(ggtree)
  library(ape)
  library(phangorn)
  library(limma)
  library(digest)
  library(factoextra)
  library(NbClust)
  library(reshape2)
  library(scales)
  library(dplyr)
  library(yarrr)
  library(plotrix)
  library(pdist)
  library(dendextend)
```


```{r, transform data}
ToxPI_results <- read.csv("Patient_Clustering_ToxPi_results.csv", header = TRUE, check.names = FALSE)

ToxPI_results <- ToxPI_results[-c(90,1), ] 


rownames(ToxPI_results) <- ToxPI_results$Name #Makes ToxPI_results row names the patient names

Names <-data.frame(ToxPI_results$Name)

ToxPI_results$`HClust Group` <- NULL
ToxPI_results$`KMeans Group` <- NULL
ToxPI_results$Source <- NULL
ToxPI_results$Name <- NULL


t_ToxPI_results <- t(ToxPI_results)

```

Plotting ToxPi profiles in R
```{r, recreate individual patient ToxPIs}

#par(mfrow = c(2,2))
#loop.vector <- 1:186
#for (i in loop.vector) {
#x<- t_ToxPI_results[,i]
      #radial.pie(x, show.grid.labels = FALSE, labels = t_ToxPI_results[,i], )
#}


#invisible(lapply(colnames(t_ToxPI_results),function(x){
     # pie(t_ToxPI_results[,x],main=x,type="l")
   # }))

```

Assembling Patient ToxPi Distance Matrix
```{r, circular dendrogram}
dist_mat <- dist(ToxPI_results, method = 'euclidean')
tupgma <- upgma(dist_mat, method = "average")

```


##Integrating code from George
```{r, echo=FALSE}
#load all desc.
clinical_info <- read.csv("ToxPi_metadata_example.csv", header= TRUE)
sortMat2 = clinical_info

#colorby property (Property is desc. col. name like disease classification)
PD = clinical_info$PD

cols3 =c("GDM" = "deepskyblue", "PRE" = "goldenrod2")

```

```{r, echo=FALSE, fig.height=12, fig.width=12, message=FALSE, warning=FALSE, include = TRUE}
#Generate Circular Dendrogram
t4 <- ggtree(tupgma, layout="circular", size=1.2)
#%<+% is a pipe symbol to combine datasets more efficiently
#merging circ. dend. w/ pvalue color assignments
t4 <- t4 %<+% clinical_info +
  geom_tippoint(aes(color=PD), alpha=0.75, size=4, shape = 15)+
  geom_text(aes(label=label, angle=angle, fontface="bold"), hjust=-0.15, size=3)+
  scale_color_manual(values =cols3) +
  theme(legend.position="bottom",legend.text=element_text(size=10))+
  theme(plot.margin = unit(c(0,0,0,0), "cm"))+
  theme(text = element_text(size = 1)) +
  geom_treescale(x = NULL, y = NULL, width = 0.5, offset = 30,
                 color = "white", linesize = 1E-100, fontsize = 1E-100)

open_tree(t4, angle = 20)

```


```{r}
to_plot2= clinical_info
rownames(to_plot2) = clinical_info$Subject.ID
```


# Other infections effects on molecular profile clustering

```{r, fig.height=12, fig.width=12, message=FALSE, warning=FALSE, include = TRUE}

colors <- c("grey80", "grey30")

#Generate Circular Dendrogram
t4 <- ggtree(tupgma, layout="circular", hjust = -.15, size=1.65)
#%<+% is a pipe symbol to combine datasets more efficiently
#merging circ. dend. w/ pvalue color assignments
t4 <- t4 %<+% clinical_info +
  geom_tippoint(aes(color=PD), alpha=0.75, size=6, shape = 18)+
  scale_color_manual(values =cols3) +
  theme(legend.position="bottom",legend.text=element_text(size=10))+
  theme(plot.margin = unit(c(0,0,0,0), "cm"))+
  theme(text = element_text(size = 1)) +
  geom_treescale(x = NULL, y = NULL, width = 0.05, offset = 30,
                 color = "white", linesize = 1E-100, fontsize = 0.01)
t4 <- gheatmap(t4, to_plot2[4], offset = .01, width = 0.12, font.size = 4, colnames = T, colnames_angle = 50)  +
  scale_fill_gradientn(colors = colors, na.value = "black")

open_tree(t4, 50) %>% rotate_tree(50)
```



# Time between sample collection and delivery on molecular profile clustering_POS

```{r, fig.height=12, fig.width=12, message=FALSE, warning=FALSE, include = TRUE}
breaks <- seq(from=min(range(-10)), to=max(range(10)), length.out=100000)
midpoint <- which.min(abs(breaks - 0))
rampCol1 <- colorRampPalette(c("#002060", "#0d78ca", "#00e8f0"))(midpoint)
rampCol2 <- colorRampPalette(c("#ff0000","#b40000","#890000", "#330000"))(100000-(midpoint))
rampCols <- c(rampCol1,rampCol2)


#Generate Circular Dendrogram
t4 <- ggtree(tupgma, layout="circular", hjust = -.15, size=1.65)
#%<+% is a pipe symbol to combine datasets more efficiently
#merging circ. dend. w/ pvalue color assignments
t4 <- t4 %<+% clinical_info +
  geom_tippoint(aes(color=PD), alpha=0.75, size=6, shape = 18)+
  scale_color_manual(values =cols3) +
  theme(legend.position="bottom",legend.text=element_text(size=10))+
  theme(plot.margin = unit(c(0,0,0,0), "cm"))+
  theme(text = element_text(size = 1)) +
  geom_treescale(x = NULL, y = NULL, width = 0.05, offset = 30,
                 color = "white", linesize = 1E-100, fontsize = 0.01)
t4 <- gheatmap(t4, to_plot2[9], offset = .12, width = 0.15, colnames = FALSE)  +
  scale_fill_gradient2(low = rampCol1, high = rampCol2, mid = "grey70", midpoint = 0)

open_tree(t4, 50) %>% rotate_tree(50)
```





